---
title: "Hands-on Exercise 1"
author: "Dr. Kam Tin Seong"
---

# Getting Started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Plotting a simple bar chart

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

# Hands-on Exercise 1

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## Hands-on Exercise 1

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

## Geom Bar

```{r}
ggplot(data=exam_data, 
       aes(x=MATHS)) +
  geom_bar()
```

## Geom Dot Plot

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)
```

## Geom Histogram

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=15)     
```

## Geom Update 1

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
    geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")
```

## Geom Update AES

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```