---
title: "Hands-on_Ex08"
---

# Choropleth Mapping with R

## Installing and loading packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Importing Data

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

### Examining mpsz content

```{r}
mpsz
```

## Importing Attribute into R

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

## Data Preparation

Prepare a data table with year 2020 values. Variables to include: PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

YOUNG: age group 0 to 4 until age group 20 to 24 ECONOMICALLY ACTIVE: age group 25-29 until age group 60-64 AGED: age group 65 and above TOTAL: all age groups, and DEPENDENCY: the ratio between YOUNG and AGED against ECONOMICALLY ACTIVE group

### Data Wrangling

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMICALLY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMICALLY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMICALLY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

### Joining the attribute data and geospatial data

This step is required to convert the values in PA and SZ fields to upper case (from upper and lower case).

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMICALLY ACTIVE` > 0)
```

This is used to join the geographical data and attribute table using planning subzone name e.g. SUBZONE_N and SZ as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

Left_join() of dplyr package is used with mpsz simple feature data frame as the left data table is to ensure that the output will be a simple features data frame.

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

## Choropleth Mapping Geospatial Data Using tmap

Two approaches can be used to prepare thematic map using tmap.

### Plot a choropleth map using qtm()

Static Map

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

Interactive Map

```{r}
tmap_mode("view")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

### Plot a choropleth map using tmap elements

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

#### Drawing a base map

In the code chunk below, tm_shape() is used to define the input data (i.e mpsz_pop2020) and tm_polygons() is used to draw the planning subzone polygons

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

### Drawing a choropleth map using tm_polygons()

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

Note:

The default interval binning used to draw the choropleth map is called “pretty”. The default colour scheme used is YlOrRd of ColorBrewer. Missing values will be shaded in grey by default.

### Drawing a choropleth map using tm_fill() and \*tm_border()\*\*

#### Using tm_full() only

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

#### Add tm_borders()

This will show the boundaries of the planning subzones.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,  
             fill_alpha = 0.1)
```

The alpha argument is used to define transparency number between 0 (totally transparent) and 1 (not transparent). By default, the alpha value of the col is used (normally 1).

There are three other arguments for tm_borders():

1)  col = border colour,
2)  lwd = border line width. The default is 1, and
3)  lty = border line type. The default is “solid”.
