---
title: "Hands-on_Ex08"
---

# Choropleth Mapping with R

## Installing and loading packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Importing Data

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

### Examining mpsz content

```{r}
mpsz
```

## Importing Attribute into R

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

## Data Preparation

Prepare a data table with year 2020 values. Variables to include: PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

YOUNG: age group 0 to 4 until age group 20 to 24 ECONOMICALLY ACTIVE: age group 25-29 until age group 60-64 AGED: age group 65 and above TOTAL: all age groups, and DEPENDENCY: the ratio between YOUNG and AGED against ECONOMICALLY ACTIVE group

### Data Wrangling

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMICALLY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMICALLY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMICALLY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

### Joining the attribute data and geospatial data

This step is required to convert the values in PA and SZ fields to upper case (from upper and lower case).

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMICALLY ACTIVE` > 0)
```

This is used to join the geographical data and attribute table using planning subzone name e.g. SUBZONE_N and SZ as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

Left_join() of dplyr package is used with mpsz simple feature data frame as the left data table is to ensure that the output will be a simple features data frame.

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

## Choropleth Mapping Geospatial Data Using tmap

Two approaches can be used to prepare thematic map using tmap.

A)  Plotting a thematic map quickly by using qtm().
B)  Plotting highly customisable thematic map by using tmap elements.

### Plot a choropleth map using qtm()

Static Map

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

Interactive Map

```{r}
tmap_mode("view")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

### Plot a choropleth map using tmap elements

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

#### Drawing a base map

In the code chunk below, tm_shape() is used to define the input data (i.e mpsz_pop2020) and tm_polygons() is used to draw the planning subzone polygons

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

### Drawing a choropleth map using tm_polygons()

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

Note:

The default interval binning used to draw the choropleth map is called “pretty”. The default colour scheme used is YlOrRd of ColorBrewer. Missing values will be shaded in grey by default.

### Drawing a choropleth map using tm_fill() and \*tm_border()\*\*

#### Using tm_full() only

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

#### Add tm_borders()

This will show the boundaries of the planning subzones.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,  
             fill_alpha = 0.1)
```

The alpha argument is used to define transparency number between 0 (totally transparent) and 1 (not transparent). By default, the alpha value of the col is used (normally 1).

There are three other arguments for tm_borders():

1)  col = border colour,
2)  lwd = border line width. The default is 1, and
3)  lty = border line type. The default is “solid”.

## Data Classification Methods of tmap

Most choropleth maps employ some methods of data classification. The purpose of data classification is to take a large number of observations and group them into data ranges or classes.

tmap provides a total ten data classification methods: fixed, sd, equal, pretty (default), quantile, kmeans, hclust, bclust, fisher, and jenks.

To define a data classification method, the style argument of tm_fill() or tm_polygons() will be used.

### Plot choropleth maps with built-in classification methods

A)  Quantile data classification using 5 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

B)  Using equal data classification method

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

C)  Quantile data classification using 2 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 2)) +
  tm_borders(fill_alpha = 0.5)
```

D)  Quantile data classification using 10 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 10)) +
  tm_borders(fill_alpha = 0.5)
```

E)  Quantile data classification using 20 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 20)) +
  tm_borders(fill_alpha = 0.5)
```

### Plot choropleth map with custom breaks

The breakpoints can be set explicitly by means of the breaks argument to the tm_fill(). In tmap(), the breaks include a minimum and maximum so to end up with n categories, n+1 elements must be specified in the breaks option (values must be in increasing order).

#### Descriptive statistics on selected variable

This is useful prior to settingthe break points

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

Based on the above, set the break point at 0.60, 0.70, 0.80, and 0.90 with a minimum and maximum of 0 and 100 respectively.

Our breaks vector is thus c(0, 0.60, 0.70, 0.80, 0.90, 1.00)

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

## Colour Schemes

### Using ColourBrewer palette

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

To reverse the colour shading, add a “-” prefix.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

## Map Layouts

Combines all map elements into a cohesive map.

### Map Legend

To change the placement, format and appearance of the legend:

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

### Map style

Classic style:

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.greens"
    ),
    col = "black",
    fill_alpha = 0.5
  ) +
  tm_layout(style = "classic")
```

### Cartographic Furniture

Add other map furniture such as compass, scale bar and grid lines:

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

To reset the default style, use the below:

```{r}
tmap_style("watercolor")
```

## Drawing Small Multiple Choropleth Maps (a.k.a facet maps)

Small multiple maps can be plotted in three ways.

### Assign Multiple Values to at least one of the Aesthetic arguments

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.scale = tm_scale_intervals(
      style = "equal",
      values = "brewer.blues"
    ),
    col_alpha = 0.5
  ) +
  tm_layout(
    legend.position = c("right", "bottom")
  ) +
  tmap_style("white")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = c("DEPENDENCY", "AGED"),
    fill.scale = list(
      tm_scale_intervals(style = "equal", values = "brewer.blues"),
      tm_scale_intervals(style = "quantile", values = "brewer.greens")
    )
  ) +
  tm_layout(
    legend.position = c("right", "bottom")
  )
```

### Define a Group-by Variable in tm_facets()

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    ),
    col = "black",
    fill_alpha = 0.5
  ) +
  tm_facets(
    by = "REGION_N",
    free.coords = TRUE
  ) +
  tm_layout(
    legend.show = FALSE,
    title.position = c("center", "center"),
    title.size = 20)
```

### Create Multiple Stand-alone Maps with tmap_arrange()

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

## Map Spatial Object Meeting a Selection Criterion

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N == "CENTRAL REGION", ]) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      hist = TRUE,
      orientation = "portrait",
      hist.height = 0.1
    ),
    col_alpha = 0.5
  ) +
  tm_layout(
    legend.outside = TRUE,
    legend.position = c("right", "bottom"),
    legend.height = 0.45,
    legend.width = 5.0,
    frame = FALSE
  )
```
