---
title: "Hands-on_Ex08a"
---

# Visualising Geospatial Point Data

Proportional symbol maps (also known as graduate symbol maps) are a class of maps that use the visual variable of size to represent differences in the magnitude of a discrete, abruptly changing phenomenon, e.g. counts of people.

Like choropleth maps, you can create classed or unclassed versions of these maps. The classed ones are known as range-graded or graduated symbols, and the unclassed are called proportional symbols, where the area of the symbols are proportional to the values of the attribute being mapped.

## Create a proportional symbol map using tmap()

Install and loading package

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Geospatial Data Wrangling

### Data Import and Prepartion

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
```

To check if data file has been imported correctly:

```{r}
list(sgpools) 
```

## Create sf data frame from an aspatial data frame

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

Notes: 1) Provide the column name of the x-coordinates first then followed by the column name of the y-coordinates. 2) The crs argument requires provision of the coordinates system in epsg format. EPSG: 3414 is Singapore SVY21 Projected Coordinate System. Other countries’ epsg codes can be found on epsg.io. 3) A new column called geometry has been added into the data frame.

To display the newly created sgpools_sf:

```{r}
list(sgpools_sf)
```

## Drawing Proportional Symbol Map

To turn on interactive mode of tmap:

```{r}
tmap_mode("view")
```

### Create an Interactive Point Symbol map

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```

### Draw a Proportional Symbol map

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

### Change Visual Attribute - Color

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

### Working with Faceted Plots

```{r}
tm_shape(sgpools_sf) + 
  tm_bubbles(fill = "OUTLET TYPE", 
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) + 
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

### Switch Viewer back to Plot mode

```{r}
tmap_mode("plot")
```
