---
title: "Hands-on_Ex10"
---

# Information Dashboard Design - R Method

## Installing and launching R packages

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
reactablefmtr, gt, gtExtras, tidyverse)
```

## Import Data Set

```{r}
coffeechain <- readr::read_rds("data/CoffeeChain.rds")
```

## Plot Bullet Chart - ggplot2 function

### Aggregate Sales & Budgets Sales (Product-level)

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`),
            .groups = "drop"
) 
```

### Plot Bullet Chart

```{r}
suppressWarnings({
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "pink",
                size = 1) +
  coord_flip()})
```

## Plot Sparklines - ggplot2 function

### Data Preparation

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

### Compute minimum, maximum and month-end sales

```{r}
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))
```

### Compute 25 and 75 quantiles

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 
                              0.25),
            quart2 = quantile(Sales, 
                              0.75)) %>%
  right_join(sales_report)
```

### Plot Spark Lines

```{r}
suppressWarnings({
  ggplot(sales_report, aes(x = Month, y = Sales)) +
    facet_grid(Product ~ ., scales = "free_y") +
    geom_ribbon(data = quarts, aes(ymin = quart1, ymax = quart2), 
                fill = 'grey90') +
    geom_line(size = 0.3) +
    geom_point(data = mins, col = 'red') +
    geom_point(data = maxs, col = 'blue') +
    geom_text(data = mins, aes(label = Sales), vjust = -1, size = 2.5) +
    geom_text(data = maxs, aes(label = Sales), vjust = 2.5, size = 2.5) +
    geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5, size = 2.5) +
    geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0, size = 2.5) +
    expand_limits(x = max(sales_report$Month) + 
                    (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
    scale_x_continuous(breaks = seq(1, 12, 1)) +
    scale_y_continuous(expand = c(0.1, 0)) +
    theme_tufte(base_size = 3) +
    theme(
      axis.title = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      strip.text = element_blank()
    )
})
```

## Static Information Dashboard Design - gt and gtExtras methods

### Using gt

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538(quiet = TRUE)
```

### Using gtExtras

-   Data Preparation

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

Code to convert the report data.frame into list columns (required).

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

## Plot Coffeechain Sales Report

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

### Add Summary Statistics

```{r}
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

### Combine data.frame

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

```{r}
sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
```

```{r}
sales_data = left_join(sales, spark)
```

### Plot Updated data.table

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

### Combine Bullet Chart and Sparklines

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup() 
```

```{r}
sales_data = sales_data %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538(quiet = TRUE)
```

## Interactive Information Dashboard Design - reactable and reactablefmtr methods

### Plot Interactive Sparklines

Prepare the list field.

```{r}
library(reactable)
library(reactablefmtr)
```

```{r}
report_spark <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales), .groups = "drop")
```

Plot the sparklines.

```{r}
reactable(
  data = report_spark,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = function(value) react_sparkline(value)
    )
  )
)
```

### Change the Page Size

Default is 10.

```{r}
reactable(
  data = report_spark,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = function(value) react_sparkline(value)
    )
  )
)
```

### Add Points & Labels

```{r}
reactable(
  data = report_spark,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = function(value) react_sparkline(
        value,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

### Add Reference Line

```{r}
reactable(
  data = report_spark, 
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = function(value) react_sparkline(
        value,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

### Add Bandline

```{r}
reactable(
  data = report_spark, 
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = function(value) react_sparkline(
        value,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

### Change from Sparkline to Sparkbar

```{r}
reactable(
  data = report_spark, 
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = function(value) react_sparkbar(
        value,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```
