---
title: "Hands-on_Ex07"
---

# Visualising and Analysing Time-oriented Data

## Installing and loading packages

Note: For version 4.5 and above, CGPfunctions cannot be installed so use alternative methods to render plots instead.

```{r}
pacman::p_load(scales, viridis, lubridate, ggthemes,
                gridExtra, readxl, knitr, data.table,
                ggHoriPlot, tidyverse, ggrepel)
```

## Data Import

```{r}
attacks <- read_csv("data/eventlog.csv")
```

## Examine Data Structure

```{r}
kable(head(attacks))
```

## Data Preparation

### Create New Fields (wkday, hour)

```{r}
make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts, 
                        tz = tz[1], 
                        quiet = TRUE)
  dt <- data.table(source_country = sc,
                   wkday = weekdays(real_times),
                   hour = hour(real_times))
  return(dt)
  }
```

### Derive Data (wkday, hour)

```{r}
wkday_levels <- c('Saturday', 'Friday', 
                  'Thursday', 'Wednesday', 
                  'Tuesday', 'Monday', 
                  'Sunday')

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp, 
                   .$source_country, 
                   .$tz)) %>% 
  ungroup() %>% 
  mutate(wkday = factor(
    wkday, levels = wkday_levels),
    hour  = factor(
      hour, levels = 0:23))
```

### Table Review

```{r}
kable(head(attacks))
```

# 1. Plotting Calendar Heatmap

## Building Calendar Heatmaps

```{r}
grouped <- attacks %>% 
  count(wkday, hour) %>% 
  ungroup() %>%
  na.omit()

ggplot(grouped, 
       aes(hour, 
           wkday, 
           fill = n)) + 
geom_tile(color = "white", 
          size = 0.1) + 
theme_tufte(base_family = "sans") + 
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
labs(x = NULL, 
     y = NULL, 
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6) )
```

## Building Multiple Calendar Heatmaps

### Derive attack by country object

```{r}
attacks_by_country <- count(
  attacks, source_country) %>%
  mutate(percent = percent(n/sum(n))) %>%
  arrange(desc(n))
```

### Preparing the tidy data frame

```{r}
top4 <- attacks_by_country$source_country[1:4]
top4_attacks <- attacks %>%
  filter(source_country %in% top4) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(
    source_country, levels = top4)) %>%
  na.omit()
```

### Plotting Multiple Calendar Heatmaps

```{r}
ggplot(top4_attacks, 
       aes(hour, 
           wkday, 
           fill = n)) + 
  geom_tile(color = "white", 
          linewidth = 0.1) + 
  theme_tufte(base_family = "sans") + 
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL, 
     title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6) )
```

# 2. Plotting Cycle Plot

## Data Import

```{r}
air <- read_excel("data/arrivals_by_air.xlsx")
```

## Data Preparation

### Create New Fields (month, year)

```{r}
air$month <- factor(month(air$`Month-Year`), 
                    levels=1:12, 
                    labels=month.abb, 
                    ordered=TRUE) 
air$year <- year(ymd(air$`Month-Year`))
```

### Extract the target country

```{r}
Vietnam <- air %>% 
  select(`Vietnam`, 
         month, 
         year) %>%
  filter(year >= 2010)
```

### Compute Average Arrivals (monthly)

```{r}
hline.data <- Vietnam %>% 
  group_by(month) %>%
  summarise(avgvalue = mean(`Vietnam`))
```

### Plotting the cycle plot

```{r}
ggplot() + 
  geom_line(data=Vietnam,
            aes(x=year, 
                y=`Vietnam`, 
                group=month), 
            colour="black") +
  geom_hline(aes(yintercept=avgvalue), 
             data=hline.data, 
             linetype=6, 
             colour="red", 
             linewidth=0.5) + 
  facet_grid(~month) +
  labs(axis.text.x = element_blank(),
       title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_tufte(base_family = "sans")
```

# 3. Plotting Slopegraph

## Data Import

```{r}
rice <- read_csv("data/rice.csv")
```

## Plotting the slopegraph - ggplot method

### Data Preparation

```{r}
df <- rice %>%
  filter(Year %in% c(1961, 1980)) %>%
  mutate(Year = factor(Year)) %>%
  group_by(Year) %>%
  mutate(rank = rank(-Yield, ties.method = "first")) %>%
  ungroup()
```

### Plotting the slopegraph

```{r}
unique(df$Country)
```

```{r}
ggplot(df, aes(x = Year, y = -rank, group = Country, color = Country)) +
  geom_line(size = 1.2, show.legend = FALSE) +
  geom_point(size = 3, show.legend = FALSE) +
  geom_text_repel(data = df %>% filter(Year == "1961"),
      aes(label = Country),
      nudge_x = -0.3, direction = "y", hjust = 1, size = 4) +
      geom_text(data = df %>% filter(Year == "1961"),
            aes(label = Yield),
            hjust = -0.3, size = 3, color = "gray40") +
  geom_text_repel(data = df %>% filter(Year == "1980"),
                  aes(label = Country),
                  nudge_x = 0.3, direction = "y", hjust = 0, size = 4) +
  geom_text(data = df %>% filter(Year == "1980"),
            aes(label = Yield),
            hjust = 1.3, size = 3, color = "gray40") +
  scale_x_discrete(expand = expansion(add = c(0.5, 0.5))) +
  scale_y_continuous(breaks = NULL) +
  labs(title = "Rice Yield of Top 11 Asian Countries",
       subtitle = "1961-1980",
       caption = "Prepared by: Cher In Teo",
       x = NULL,
       y = NULL) +
  theme_minimal(base_size = 14) +
  theme(panel.grid = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.caption = element_text(size = 10, hjust = 1),
        legend.position = "none")
```
