---
title: "Hands-on_Ex08"
---

# Data Classification Methods of tmap

Most choropleth maps employ some methods of data classification. The purpose of data classification is to take a large number of observations and group them into data ranges or classes.

tmap provides a total ten data classification methods: fixed, sd, equal, pretty (default), quantile, kmeans, hclust, bclust, fisher, and jenks.

To define a data classification method, the style argument of tm_fill() or tm_polygons() will be used.

## Plot choropleth maps with built-in classification methods

A)  Quantile data classification using 5 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

B)  Using equal data classification method

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

C)  Quantile data classification using 2 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 2)) +
  tm_borders(fill_alpha = 0.5)
```

D)  Quantile data classification using 10 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 10)) +
  tm_borders(fill_alpha = 0.5)
```

E)  Quantile data classification using 20 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 20)) +
  tm_borders(fill_alpha = 0.5)
```

## Plot choropleth map with custom breaks

The breakpoints can be set explicitly by means of the breaks argument to the tm_fill(). In tmap(), the breaks include a minimum and maximum so to end up with n categories, n+1 elements must be specified in the breaks option (values must be in increasing order).

### Descriptive statistics on selected variable

This is useful prior to settingthe break points

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

Based on the above, set the break point at 0.60, 0.70, 0.80, and 0.90 with a minimum and maximum of 0 and 100 respectively.

Our breaks vector is thus c(0, 0.60, 0.70, 0.80, 0.90, 1.00)

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

## Colour Schemes

### Using ColourBrewer palette

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

To reverse the colour shading, add a “-” prefix.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

## Map Layouts

Combines all map elements into a cohesive map.

### Map Legend

To change the placement, format and appearance of the legend:

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

### Map style

Classic style:

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.greens"
    ),
    col = "black",
    fill_alpha = 0.5
  ) +
  tm_layout(style = "classic")
```

### Cartographic Furniture

Add other map furniture such as compass, scale bar and grid lines:

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

To reset the default style, use the below:

```{r}
tmap_style("watercolor")
```

## Drawing Small Multiple Choropleth Maps (a.k.a facet maps)

Small multiple maps can be plotted in three ways.

### Assign Multiple Values to at least one of the Aesthetic arguments

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.scale = tm_scale_intervals(
      style = "equal",
      values = "brewer.blues"
    ),
    col_alpha = 0.5
  ) +
  tm_layout(
    legend.position = c("right", "bottom")
  ) +
  tmap_style("white")
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = c("DEPENDENCY", "AGED"),
    fill.scale = list(
      tm_scale_intervals(style = "equal", values = "brewer.blues"),
      tm_scale_intervals(style = "quantile", values = "brewer.greens")
    )
  ) +
  tm_layout(
    legend.position = c("right", "bottom")
  )
```

### Define a Group-by Variable in tm_facets()

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    ),
    col = "black",
    fill_alpha = 0.5
  ) +
  tm_facets(
    by = "REGION_N",
    free.coords = TRUE
  ) +
  tm_layout(
    legend.show = FALSE,
    title.position = c("center", "center"),
    title.size = 20)
```

### Create Multiple Stand-alone Maps with tmap_arrange()

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

## Map Spatial Object Meeting a Selection Criterion

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N == "CENTRAL REGION", ]) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(
      hist = TRUE,
      orientation = "portrait",
      hist.height = 0.1
    ),
    col_alpha = 0.5
  ) +
  tm_layout(
    legend.outside = TRUE,
    legend.position = c("right", "bottom"),
    legend.height = 0.45,
    legend.width = 5.0,
    frame = FALSE
  )
```
