---
title: "Hands-on_Ex08b"
---

# Analytical Mapping using R

## Installing and Loading Packages

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

## Importing data

```{r}
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

## Basic Choropleth Mapping

### Visualising Distribution of Non-functional Water Point

Plot a choropleth map showing the distribution of non-function water point by LGA.

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
             fill.scale = tm_scale_intervals(
               style = "equal",
               n = 10,
               values = "brewer.blues"),
             fill.legend = tm_legend(
               position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of functional water point by LGAs")
```

```{r}
p2 <- tm_shape(NGA_wp) + 
  tm_polygons(fill = "total_wp", 
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1, 
             fill_alpha = 1) + 
  tm_title("Distribution of total  water point by LGAs")
```

```{r}
tmap_arrange(p2, p1, nrow = 1)
```

## Choropleth Map for Rates

### Deriving Proportion of Functional Water Points and Non-Functional Water Points

Mutate() from dplyr package is used to derive two fields, namely pct_functional and pct_nonfunctional.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

### Plotting Map of Rate

```{r}
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) + 
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Rate map of functional water point by LGAs")
```

## Extreme Value Maps

Extreme value maps are variations of common choropleth maps where the classification is designed to highlight extreme values at the lower and upper end of the scale, with the goal of identifying outliers.

These maps were developed in the spirit of spatializing EDA, i.e., adding spatial features to commonly used approaches in non-spatial EDA (Anselin 1994).

### Percentile Map

The percentile map is a special type of quantile map with six specific categories: 0-1%,1-10%, 10-50%,50-90%,90-99%, and 99-100%. The corresponding breakpoints can be derived by means of the base R quantile command, passing an explicit vector of cumulative probabilities as c(0,.01,.1,.5,.9,.99,1). Note that the begin and endpoint need to be included.

#### Data Preparation

1)  Exclude records with NA by using the code chunk below.

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

2)  Creating customised classification and extracting values

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

Note: Many base R functions cannot deal with the geometry so st_set_geometry(NULL) is used to drop geometry field.

## Writing Functions

Advantages: 1) Can give a function an evocative name that makes code easier to understand. 2) As requirements change, only need to update code in one place. 3) Eliminate the chance of making incidental mistakes during copy and paste (i.e. updating a variable name in one place, but not in another).

### Create the get.var function

Write an R function as shown below to extract a variable (i.e. wp_nonfunctional) as a vector out of an sf data.frame.

-   arguments: vname: variable name (as character, in quotes) df: name of sf data frame

-   returns: v: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% 
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

### A Percentile Mapping Function

Write a percentile mapping function:

```{r}
percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {

  percent <- c(0, .01, .1, .5, .9, .99, 1)
  var <- df[[vnam]]
  bperc <- quantile(var, percent, na.rm = TRUE)

  tm_shape(df) +
    tm_polygons(
      fill = vnam,
      fill.scale = tm_scale_intervals(
        breaks = bperc,
        values = "brewer.blues",
        labels = c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%")
      ),
      fill.legend = tm_legend(title = legtitle),
      col = "black"
    ) +
    tm_title(mtitle) +
    tm_layout(title.position = c("right","bottom"))
}
```

### Test the Percentile Mapping Function

```{r}
percentmap("total_wp", NGA_wp)
```

## Box map

A box map is an augmented quartile map, with an additional lower and upper category. When there are lower outliers, then the starting point for the breaks is the minimum value, and the second break is the lower fence.

When there are no lower outliers, then the starting point for the breaks will be the lower fence, and the second break is the minimum value (there will be no observations that fall in the interval between the lower fence and the minimum value).

```{r}
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

The break points for the box map vary depending on whether lower or upper outliers are present.

## Create the Boxbreaks function

The code chunk below is an R function that creating break points for a box map.

-   arguments: v: vector with observations mult: multiplier for IQR (default 1.5)

-   returns: bb: vector with 7 break points compute quartile and fences

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

### Create the get.var function

-   arguments: vname: variable name (as character, in quotes) df: name of sf data frame

-   returns: v: vector with values (without a column name)

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### Test the newly created function

```{r}
var <- get.var("wp_nonfunctional", NGA_wp) 
boxbreaks(var)
```

## Boxmap Function

To create a box map:

-   arguments: vnam: variable name (as character, in quotes) df: simple features polygon layer legtitle: legend title mtitle: map title mult: multiplier for IQR -

-   returns: a tmap-element (plots a map)

```{r}
boxmap <- function(vnam, df,
                   legtitle = NA,
                   mtitle = "Box Map",
                   mult = 1.5){

  var <- df[[vnam]]
  bb <- boxbreaks(var, mult = mult)

  tm_shape(df) +
    tm_fill(
      col = vnam,
      title = legtitle,
      breaks = bb,
      palette = "Blues",
      labels = c(
        "Lower outlier",
        "< 25%",
        "25% - 50%",
        "50% - 75%",
        "> 75%",
        "Upper outlier"
      )
    ) +
    tm_borders() +
    tm_layout(
      main.title = mtitle,
      title.position = c("left", "top")
    )
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
